---
title: "CS3244 Linear Regression Pipeline"
author: "A0272756U"
date: "2026-04-10"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(caret)
library(dplyr)
library(rlang)
library(tidyr)
```

## 1. Data Preperation

```{r}
# Load data
data <- read.csv("C:/Users/Study/OneDrive - National University of Singapore/Desktop/Y3S2/CS3244/CS3244_group-3_ResalePriceModeling/data/hdb_resale_final_modeling_dataset_2015_2025.csv")

# Clean data for modeling
data_modelling <- data %>%
  mutate(across(where(is.character), as.factor)) %>%
  drop_na() %>%
  dplyr::select(-resale_price, -region, -town, -flat_model_raw) # Exclude unused high-cardinality/redundant columns
```

## 2. Baseline Establishment

**Action:** Build a baseline model using only original structural features.

**Metrics:** Adjusted $R^2$ and exact-dollar RMSE.

```{r}
# Fit Baseline Model
baseline_model <- lm(log_resale_price ~ floor_area_sqm + remaining_lease_years + storey_mid + flat_type + flat_model, data = data_modelling)

# Extract Adjusted R-Squared
adj_r_sq_base <- summary(baseline_model)$adj.r.squared

# Calculate exact-dollar metrics
preds_actual_base <- exp(predict(baseline_model, data_modelling))
actuals_base <- exp(data_modelling$log_resale_price)

rmse_base <- sqrt(mean((actuals_base - preds_actual_base)^2))
# NEW: Mean Bias Error (Actual - Predicted)
mbe_base <- mean(actuals_base - preds_actual_base)

cat(sprintf("Baseline Adjusted R-squared: %.4f\n", adj_r_sq_base))
cat(sprintf("Baseline RMSE: $%.2f\n", rmse_base))
cat(sprintf("Baseline MBE: $%.2f\n", mbe_base))
```

## 3. Model Adequacy Checks

```{r}
# 1. Correlation for Numerical Features Only
numeric_data <- data_modelling %>% 
  dplyr::select(where(is.numeric))

# Visual correlation matrix (using GGally)
ggcorr(numeric_data, label = TRUE, label_size = 3, hjust = 1, layout.exp = 2)

# 2. Multicollinearity (Handles both Categorical and Numerical)
# Fit a base linear model. We exclude log_resale_price to prevent target leakage.
baseline_model <- lm(log_resale_price ~ . , data = data_modelling)

# Calculate GVIF using the car package
# Features with GVIF^(1/(2*Df)) > 2 to 2.5 indicate high multicollinearity
vif_results <- car::vif(baseline_model)
print(vif_results)
```

## Data cleaning for modelling

```{r}
data_modelling <- data %>%
  
  # 1. Convert character strings to factors (critical for regression dummy variables)
  
  mutate(across(where(is.character), as.factor)) %>%

  # 2. Drop rows with NA values in the distance columns
  drop_na() %>%
  dplyr::select(-resale_price, -region, -town, -flat_model_raw)
  
# Check the structure to ensure factor levels are set correctly
glimpse(data_modelling)
```

## Checking model assumptions

(Before training)

1.  Check that response has a normal distribution (qqplot + qqnorm)
2.  check GVIF for multicollinearity

### 3.1 Normality of resale price distribution

```{r resale histogram plot}
p_orig <- ggplot(data, aes(x = resale_price)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(title = "Original resale price", x = "Price (SGD)", y = "Count") +
  theme_minimal()

p_log10 <- ggplot(data, aes(x = log(resale_price))) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  labs(title = "Log Transformed price", x = "log(Price)", y = "Count") +
  theme_minimal()

p_orig + p_log10
```

### 3.2 Checking for multicollinearity

#### 3.2.1 Correlation matrix (Numerical vs Numerical)

```{r}
# 1. Correlation for Numerical Features Only
numeric_data <- data_modelling %>% 
  dplyr::select(where(is.numeric))

# Visual correlation matrix (using GGally)
ggcorr(numeric_data, label = TRUE, label_size = 3, hjust = 1, layout.exp = 2)

# 2. Multicollinearity (Handles both Categorical and Numerical)
# Fit a base linear model. We exclude log_resale_price to prevent target leakage.
baseline_model <- lm(log_resale_price ~ . , data = data_modelling)

```

#### 3.2.2 GVIF (Generalised multicollinearity check)

```{r}
# Calculate GVIF using the car package
# Features with GVIF^(1/(2*Df)) > 2 to 2.5 indicate high multicollinearity
vif_results <- car::vif(baseline_model)
print(vif_results)
```

#### 3.2.3 Continuous variable correlation check with $R^2$

```{r}
# Define your continuous target variable
cont_target <- "floor_area_sqm"

# Define all the categorical features you want to test against it
cat_features <- c("flat_type", "flat_model", "region_URA", "mature_estate")

cat(sprintf("\n--- Collinearity Diagnostics for '%s' ---\n", cont_target))
cat(sprintf("%-15s | %-15s | %-10s\n", "Feature", "Variance Expl.", "VIF Equivalent"))
cat(paste(rep("-", 48), collapse=""), "\n")

for (col in cat_features) {
  # Dynamically construct the formula
  formula_str <- paste(cont_target, "~", col)
  
  # Fit the model and suppress any singular fit warnings
  model <- suppressWarnings(lm(as.formula(formula_str), data = data_modelling))
  
  # Extract the R-squared value
  r_sq <- summary(model)$r.squared
  
  # Calculate VIF (handling the edge case of perfect collinearity preventing division by zero)
  if (r_sq >= 0.999) {
    vif_val <- Inf
  } else {
    vif_val <- 1 / (1 - r_sq)
  }
  
  # Output the formatted row
  cat(sprintf("%-15s | %.3f (%.1f%%)   | %.2f\n", col, r_sq, r_sq * 100, vif_val))
}
```

#### 3.2.4 Categorical variable correlation check

```{r}
# 1. Define Cramer's V (Categorical vs Categorical) with Bias Correction
cramers_v <- function(x, y) {
  # Remove missing values to prevent table errors
  valid_idx <- complete.cases(x, y)
  x <- as.factor(x[valid_idx])
  y <- as.factor(y[valid_idx])
  
  tbl <- table(x, y)
  
  # Suppress warnings for expected counts (safe due to massive N)
  chi2 <- suppressWarnings(chisq.test(tbl, correct = FALSE)$statistic)
  
  n <- sum(tbl)
  r <- nrow(tbl)
  k <- ncol(tbl)
  
  # Apply bias correction
  phi2 <- chi2 / n
  phi2corr <- max(0, phi2 - ((k - 1) * (r - 1)) / (n - 1))
  rcorr <- r - ((r - 1)^2) / (n - 1)
  kcorr <- k - ((k - 1)^2) / (n - 1)
  
  cv <- sqrt(phi2corr / min((kcorr - 1), (rcorr - 1)))
  return(as.numeric(cv))
}

# Run tests for categorical features
cat_features <- c("mature_estate", "flat_type")
cat("\n--- Cramer's V (Categorical vs region_URA) ---\n")

for (col in cat_features) {
  # CHANGED: Using data_modelling instead of df
  score <- cramers_v(data$region_URA, data[[col]])
  cat(sprintf("region_URA vs %s: %.3f\n", col, score))
}


# 2. Define Eta-Squared (Categorical vs Continuous)
eta_squared <- function(cat_col_name, cont_col_name, data) {
  # Create formula: continuous ~ categorical
  formula_str <- paste(cont_col_name, "~", cat_col_name)
  
  # Fit ANOVA model
  model <- aov(as.formula(formula_str), data = data)
  aov_table <- summary(model)[[1]]
  
  # Calculate Eta-Squared: SS(Between) / SS(Total)
  ss_between <- aov_table$`Sum Sq`[1]
  ss_residuals <- aov_table$`Sum Sq`[2]
  ss_total <- ss_between + ss_residuals
  
  eta2 <- ss_between / ss_total
  return(eta2)
}

# Run tests for continuous features (The suspected spatial conflict)
cont_features <- c("dist_to_cbd_m", "mrt_walking_distance", "floor_area_sqm")
cat("\n--- Eta-Squared (Continuous vs region_URA) ---\n")

for (col in cont_features) {
  # CHANGED: Passing data_modelling instead of df
  score <- eta_squared("region_URA", col, data)
  cat(sprintf("region_URA vs %s: %.3f\n", col, score))
}
```

**Craimer V thesholds:**

0.00 to 0.29 (Safe Zone)

0.30 to 0.59 (Warning Zone - Moderate Collinearity)

0.60 to 1.00 (Danger Zone - Severe Collinearity)

2\. Eta-Squared ($\eta^2$) (Categorical vs. Continuous)

Eta-squared represents the exact percentage of variance explained (e.g., 0.14 = 14%).

-   **\~ 0.01 (1%):** Small effect

-   **\~ 0.06 (6%):** Medium effect

-   **\> 0.14 (14%+):** Large effect

verdict:

Remove days_since_start (due to high correlation in correlation matrix)

Remove flat_type (due to high VIF with floor_area_sqm)

Remove region_URA due to high craimer V and eta squared score

## 4. Model building 

### 4.1 Initial model

```{r}
full_model = lm(log_resale_price ~ mature_estate + flat_model + floor_area_sqm + storey_mid + remaining_lease_years + mrt_walking_distance + dist_to_cbd_m + dist_nearest_park_m + dist_nearest_hawker_centre_m + sora_monthly_avg + Per_Capita_GDP, data = data_modelling)

stepAIC(full_model, direction="backward", k = log(258649))  # for backward selection

summary(full_model)
anova(full_model)

vif_results <- car::vif(full_model)
print(vif_results)    
```

### 4.2 Final model

```{r}
final_model = lm(log_resale_price ~ Per_Capita_GDP + floor_area_sqm + storey_mid + remaining_lease_years + dist_to_cbd_m + sora_monthly_avg + mature_estate, data = data_modelling)


stepAIC(final_model, direction="backward", k = log(258649))  # for backward selection

summary(final_model)
anova(final_model)

vif_results <- car::vif(final_model)
print(vif_results)
```

Because your dependent variable is log-transformed, the exact percentage change in resale price for a 1-unit increase in the predictor is calculated using this formula:

**Percentage Change** = $(e^{\beta} - 1) \times 100$

| Feature | Original Estimate ($\beta$) | Exact Percentage Change | Interpretation (for a 1-unit increase) |
|:-----------------|:-----------------|:-----------------|:-----------------|
| **mature_estateMature** | 0.1154 | **+12.23%** | Mature estates are 12.23% more expensive than non-mature estates. |
| **floor_area_sqm** | 0.009683 | **+0.97%** | Each additional square meter increases price by 0.97%. |
| **storey_mid** | 0.008783 | **+0.88%** | Each higher floor increases price by 0.88%. |
| **remaining_lease_years** | 0.009374 | **+0.94%** | Each extra year of remaining lease increases price by 0.94%. |
| **dist_to_cbd_m** | -0.00002349 | **-0.00235%** | Each meter further from the CBD *decreases* price by 0.00235% (or -2.35% per kilometer). |
| **sora_monthly_avg** | 0.04535 | **+4.64%** | A 1-unit increase in SORA increases price by 4.64%. |
| **Per_Capita_GDP** | 0.00001688 | **+0.00169%** | A \$1 increase in GDP per capita increases price by 0.00169% (or +1.69% per \$1000 increase). |

*Note: For very small coefficients (like distance to CBD or GDP), the percentage change is nearly identical to just multiplying the estimate by 100. However, for larger coefficients (like mature estate), the exponential formula is strictly required for accuracy.*

#Evalaluating final model
```{r}
# Calculate exact-dollar metrics for Final Model
preds_actual_final <- exp(predict(final_model, data_modelling))
actuals_final <- exp(data_modelling$log_resale_price)

rmse_final <- sqrt(mean((actuals_final - preds_actual_final)^2))
# NEW: Mean Bias Error for Final Model
mbe_final <- mean(actuals_final - preds_actual_final)

cat(sprintf("Final Model RMSE: $%.2f\n", rmse_final))
cat(sprintf("Final Model MBE: $%.2f\n", mbe_final))

# Quick comparison logic
bias_direction <- if(mbe_final > 0) "underpredicting" else "overpredicting"
cat(sprintf("On average, the final model is %s prices by $%.2f\n", bias_direction, abs(mbe_final)))
```

#calculate CV-RMSE
```{r}
# Function for calculating CVRMSE 
calculate_cvrmse <- function(actuals, predicts) {
  rmse <- sqrt(mean((actuals - predicts)^2))
  mean_actual <- mean(actuals)
  cvrmse <- (rmse / mean_actual) * 100 
  return(cvrmse)
}

# 1. CVRMSE for Baseline Model
actuals_base <- exp(data_modelling$log_resale_price)
preds_base   <- exp(predict(baseline_model, data_modelling))
cvrmse_base <- calculate_cvrmse(actuals_base, preds_base)

# 2. CVRMSE for Final Model
actuals_final <- exp(data_modelling$log_resale_price)
preds_final   <- exp(predict(test_model, data_modelling))
cvrmse_final <- calculate_cvrmse(actuals_final, preds_final)

#Output Results 
cat(sprintf("Baseline CVRMSE: %.2f%%\n", cvrmse_base))
cat(sprintf("Final Model CVRMSE: %.2f%%\n", cvrmse_final))
```
## 5. Model Adequacy Checks

### 5.1 Checking validity of dimensionality reduction techniques

```{r}
# 1. Prepare the evaluation dataset
# Drop NAs first so the number of rows fed into the model perfectly matches the predictions.
# We DO NOT select/drop any columns here, so 'town' and 'flat_model_raw' are preserved.
data_eval <- data %>%
  mutate(across(where(is.character), as.factor)) %>%
  drop_na(dist_to_cbd_m, dist_nearest_park_m, dist_nearest_hawker_centre_m)

# 2. Build the model on this unified dataset
# (Ensure this formula matches your exact finalized features)
test_model_eval <- lm(log_resale_price ~ mature_estate +
                        floor_area_sqm + storey_mid + remaining_lease_years + 
                        mrt_walking_distance + dist_to_cbd_m + Per_Capita_GDP, 
                      data = data_eval)

# 3. Append predictions directly to the dataset
data_eval <- data_eval %>%
  mutate(
    log_predictions = predict(test_model_eval, newdata = data_eval),
    predicted_price = exp(log_predictions) # Convert back to actual dollars for evaluation
  )

# 4. Define the evaluation function
evaluate_group_performance <- function(df, group_col, actual_col = "resale_price", pred_col = "predicted_price") {
  df %>%
    group_by(!!sym(group_col)) %>%
    summarise(
      N = n(),
      RMSE = sqrt(mean((!!sym(actual_col) - !!sym(pred_col))^2)),
      MAE = mean(abs(!!sym(actual_col) - !!sym(pred_col))),
      Mean_Bias = mean(!!sym(actual_col) - !!sym(pred_col)) # Positive = Underpredicted
    ) %>%
    arrange(desc(RMSE)) %>%
    as.data.frame()
}

# 5. Run the spatial and structural evaluations
cat("\n--- Top 10 Highest RMSE by Town ---\n")
town_diagnostics <- evaluate_group_performance(data_eval, "town")
print(town_diagnostics)

cat("\n--- Top 10 Highest RMSE by Raw Flat Model ---\n")
flat_model_diagnostics <- evaluate_group_performance(data_eval, "flat_model_raw")
print(flat_model_diagnostics)
```

### 5.2 Checking constant variance and mean 0 assumptions

```{r}
### 4.3 Model Adequacy Checking
# Extract standardized residuals
SR_final <- rstandard(final_model)

# 1. Normality of Residuals (QQ Plot)
# Look for adherence to the red diagonal line. Due to large N, expect some tail deviation.
qqnorm(SR_final, main = "QQ Plot of Standardized Residuals", pch = 20, col = rgb(0,0,0,0.05))
qqline(SR_final, col = "red", lwd = 2)

# 2. Linearity & Constant Variance (Residuals vs Fitted)
# Look for a random horizontal band around 0 without funnel shapes (heteroscedasticity).
plot(fitted(final_model), SR_final, 
     main = "Standardized Residuals vs Fitted Values", 
     xlab = "Fitted log_resale_price", 
     ylab = "Standardized Residuals", 
     pch = 20, col = rgb(0,0,0,0.05))
abline(h = 0, col = "red", lwd = 2)
```

```{r}
# 3. Residuals vs Predictors 
# Look for random scatter. Non-random patterns (like curves) indicate missing higher-order terms.
par(mfrow = c(2, 3)) # Set up a 2x3 grid for the 6 continuous variables

plot(data_modelling$floor_area_sqm, SR_final, 
     main = "Residuals vs Floor Area", xlab = "Floor Area (sqm)", 
     ylab = "Standardized Residuals", pch = 20, col = rgb(0,0,0,0.05))
abline(h = 0, col = "red", lwd = 2)

plot(data_modelling$dist_to_cbd_m, SR_final, 
     main = "Residuals vs Distance to CBD", xlab = "Distance to CBD (m)", 
     ylab = "Standardized Residuals", pch = 20, col = rgb(0,0,0,0.05))
abline(h = 0, col = "red", lwd = 2)

plot(data_modelling$storey_mid, SR_final, 
     main = "Residuals vs Storey Midpoint", xlab = "Storey (Mid)", 
     ylab = "Standardized Residuals", pch = 20, col = rgb(0,0,0,0.05))
abline(h = 0, col = "red", lwd = 2)

plot(data_modelling$remaining_lease_years, SR_final, 
     main = "Residuals vs Remaining Lease", xlab = "Lease Years Remaining", 
     ylab = "Standardized Residuals", pch = 20, col = rgb(0,0,0,0.05))
abline(h = 0, col = "red", lwd = 2)

plot(data_modelling$Per_Capita_GDP, SR_final, 
     main = "Residuals vs Per Capita GDP", xlab = "Per Capita GDP", 
     ylab = "Standardized Residuals", pch = 20, col = rgb(0,0,0,0.05))
abline(h = 0, col = "red", lwd = 2)

plot(data_modelling$sora_monthly_avg, SR_final, 
     main = "Residuals vs SORA", xlab = "SORA Monthly Avg", 
     ylab = "Standardized Residuals", pch = 20, col = rgb(0,0,0,0.05))
abline(h = 0, col = "red", lwd = 2)

par(mfrow = c(1, 1)) # Reset plot window back to 1x1
```
